---
title: 1. Introduction
author: Matthew E. Dardet
date: "2024-09-05"
slug: 1-introduction
categories: []
tags: []
---

$$
Y_{2024_i} = 0.75 \times Y_{2020\_voteshare_i} + 0.25 \times Y_{2016\_voteshare_i}
$$

```{r, echo = FALSE, message = FALSE, warning = FALSE}
####----------------------------------------------------------#
#### Preamble
####----------------------------------------------------------#

# Load libraries.
## install via `install.packages("name")`
library(ggplot2)
library(maps)
library(tidyverse)

####----------------------------------------------------------#
#### Read and clean presidential popular vote.
####----------------------------------------------------------#

# Read presidential popular vote. 
d_popvote <- read_csv("data/popvote_1948-2020.csv")

# Pivot data to wide format with party names as columns and two-party vote share as values.
d_popvote_wide <- d_popvote |>
  select(year, party, pv2p) |>
  pivot_wider(names_from = party, values_from = pv2p)

# Modify winner column to show "D" if Democrats win and "R" if Republicans win. 
d_popvote_wide <- d_popvote_wide |> 
  mutate(winner = case_when(democrat > republican ~ "D",
                            TRUE ~ "R"))
```

``````{r, echo = FALSE, message = FALSE, warning = FALSE}
####----------------------------------------------------------#
#### Visualize trends in national presidential popular vote. 
####----------------------------------------------------------#

# Example: bad histogram; unpleasant bin width and fill!
plt.1 <- ggplot(d_popvote, aes(x = pv2p)) + 
    geom_histogram() + 
    labs(x = "Two-Party Vote Share", 
         y = "Count", 
         title = "Distribution of Two-Party Vote Share (1948-2020)")

# Example: better histogram.
plt.2 <- ggplot(d_popvote, aes(x = pv2p)) + 
  geom_histogram(bins = 10, color = "black", fill = "goldenrod2") + 
  theme_bw() +
  labs(x = "Two-Party Vote Share", 
       y = "Count", 
       title = "Distribution of Two-Party Vote Share (1948-2020)")

# Example: barplot (+ custom colors).
plt.3 <- ggplot(d_popvote, aes(x = year, y = pv2p, fill = party)) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = c("dodgerblue4", "firebrick1")) 

# Example: lineplot (+ custom colors + nicer theme).
plt.4 <- ggplot(d_popvote, aes(x = year, y = pv2p, colour = party)) +
    geom_line() +
    scale_color_manual(values = c("dodgerblue4", "firebrick1")) + 
    theme_bw()

# Example: bad line plot; opaque background, "too much ink", no legend, small font. Yuck! 
plt.5 <- ggplot(d_popvote, aes(x = year, y = pv2p, colour = party)) +
    geom_line(stat = "identity") + 
    theme_dark() +
    theme(legend.position = "none", axis.title = element_text(size = 5))

# Example: good line plot: high contrast, "minimal ink", legend, detailed x-ticks, larger font.
plt.6 <- ggplot(d_popvote, aes(x = year, y = pv2p, colour = party)) +
    geom_line(stat = "identity") + 
    scale_x_continuous(breaks = seq(1948, 2020, 4)) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45))

# Example: excellent plot; "pretty" customized theme.
my_pretty_theme <- theme_bw() + 
    theme(panel.border = element_blank(),
          plot.title = element_text(size = 15, hjust = 0.5), 
          axis.text.x = element_text(angle = 45, hjust = 1),
          axis.text = element_text(size = 12),
          strip.text = element_text(size = 18),
          axis.line = element_line(colour = "black"),
          legend.position = "top",
          legend.text = element_text(size = 12))

ggplot(d_popvote, aes(x = year, y = pv2p, colour = party)) +
    geom_line(stat = "identity") +
    scale_color_manual(values = c("dodgerblue4", "firebrick1"), name = "") +
    xlab("") + ## no need to label an obvious axis
    ylab("Two-Party National Popular Vote (%)") +
    ggtitle("Presidential Vote Share (1948-2020)") + 
    scale_x_continuous(breaks = seq(1948, 2020, 4)) +
    my_pretty_theme
```




